### 2009-12-08 docs/marcello-dissertation/thesis/chapter8-experimental-results-correct-behavior-and-performance.tex
First Draft of the dissertation peer reviewed with Kleber.
git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@680 559edde2-a4d4-44e0-96fc-4035e40d7396

The simulation of both the single and distributed system used both a Mac OS 10
Snow Leopard 64bit, with a total of 8GB of physical memory. However, in order
to simulate different shards in different machines, the use of a VirtualBox
\cite{virtualization} provided the use of guest operating systems to run and
act as the mongoDB shards. In this way, two different versions of the Ubuntu
Linux were used: one 8.04 32bits and another 9.04 64bits (see section
\ref{sec:experiments-difficulties}).

@@

The first set of measurements collected were the ones regarding the use case
R1, shown in Listing \ref{file:dsp-mongo-service}, defined as the insertion of
instances of the collected data from the sensor devices using Java. The use of
the mongoDB Java driver, which provides the methods to access the database
system and the collection. In this way, the construction of the keys and
respective values were done using a Hash-like object ``BasicDBObject'' such as
the ones shown in the method ``buildKeySegment''. Similarly, the creation of
database indexes is performed using the same approach, as shown in the Java
method ``setupMetadataIndexes''. Finally, the insertion of the documents is
performed on the service method call ``insertPersistentUnitMessageContents'',
which delimits the bulk insert using the transaction management mechanism over
the collection. In this way, different insert times were collected from the
logs of the different experiments as summarized on Table
\ref{tab:experiment-insert-avarage}.

@@

\begin{table}
    \begin{center}
        \begin{tabular}{|p{70pt}|p{100pt}|p{250pt}|p{100pt}|}\hline
        \textbf{Scenario Type} & \textbf{Language} & \textbf{Method Call} & \textbf{Use Case}\\\hline 
        \textbf{Insert} & Java & db.SondeDataContainer.insert() & R0\\\hline
        \textbf{Retrive} & Javascript & db.SondeDataContainer.find() & R1, R2, R3\\\hline 
        \textbf{Update} & Javascript & db.SondeDataContainer.update() & U1\\\hline 
        \textbf{Delete} & Javascript & db.SondeDataContainer.remove() & D1\\\hline
        \textbf{Export} & Python & db.SondeDataContainer.find() & R4\\\hline
        \end{tabular}
        \caption{mongoDB method calls used for the use cases}
    \end{center}
    \label{tab:experiment-scenarios-mongo-methods}
\end{table}

@@

\subsection{User Experience}

Different methods of data access were used after the randomly generated data
was inserted into mongoDB: using the shell commands mongo and ``mongoexport'', 
command-line execution developed in Python, through the REST Web Services API,
and using a web application, written in the Ruby Programming Language
\ref{ruby}, providing the view over the collected data using a web browser.
More details about the execution of these commands during the experiments is
shown in section \ref{sec:mongodb-user-experience}.

@@

While the access to the database using the shell command gives full access to 
the CRUD operations, the access to the data was also performed using the REST
Web Services API, as shown in Listing \ref{cmd:mongo-rest-request}. As one can
see, the IP address and the default port number of the mongoDB REST middleware
are provided. Then, the REST operation uses the name of the database and
the collection to correlate with the ones in the database system. In this way,
querying the collections from mongoDB can be performed using the HTTP Request
Parameter variables, by adding the prefix ``filter_'' before the name of the
key of the document. The given example shows the query performed in the
collected data for the key ``observation.Conductivity'' with the value
``104.5''. Similarly, the HTTP Request variable ``limit'' is used to reduce
the number of documents returned by the query. Finally, the HTTP Response from
mongoDB returns the result using the JSON format, similar to the ones printed
in the command-shell.

@@

In order to verify

@@

The implementation of the data persistence for NetBEAMS follows the
specifications explained in chapter 4. The purpose of sensor network for
NetBEAMS is to suport data archival from any collected data, and for this
reason, mongoDB was selected after an empirical analysis described in chapter
5. In this way, the requirements described in chapter 4 were expressed as
examples of use cases in section \ref{sec:exp-scenarios}.

All the collected data from sensors are stored in a
given single server, which characterizes an External Storage device

the data. The process to describe the data is done with prior knowledge of the
sensor device data properties when choosing the relational data model, because
of the process of database normalization \cite{db-normalization}, which takes
place after the the entities have been identified. Consequently, the
introduction of a new sensor device might potentially require changes on the
database schema structure conceived before, as the process of database
normalization must follow.

@@

Any database system covers the purpose of the sensor data for NetBEAMS, which
is Data Archival, along with the type of storage for the data as an External
Storage. After selecting mongoDB, the development of the persistence layer met
the requirements of archiving data after inserting the the randomly generated
data that represents the data for the YSI sonde. As it is shown in section
\ref{sec:exp-measurements}, the collected data is archived in files using the
BSON format \cite{bson}, a binary representation used to store instances of
the document described in Listing \ref{file:mongodb-ysi-data-format}. Along
with the purpose of data archival, the use of mongoDB also met the
specification of an external storage since the data persisted in the file
system, as well as Data-Centric for the proposal to use mongoDB distributed
shards.

SF-BEAMS uses the Tabular Data model to describe the collected data from the
sensor devices. As discussed in chapter 5, the decision to use mongoDB is the
fact that the literature review does not mention the use of the Key-Value-Pair
Data model. For this reason, since mongoDB implements a variation of the 
key-value data model based on the JSON \cite{json} data format, the Data
Model taxonomy should be updated, as shown in Figure
\ref{fig:taxonomy-data-model-modified}.

\begin{figure}[h]
  \centering
  \includegraphics[scale=0.5]{../diagrams/taxonomy-data-model-modified}
  \caption{Data Model Taxonomy}
  \label{fig:taxonomy-data-model-modified}
\end{figure}

Another aspect of the implementation is that provides APIs in different
languages for the data creation and extraction processes. In this way, the
implementation of the persistence model captures the instances of any known
data collected from NetBEAMS and saves it using the document pattern as shown
in Listing \ref{file:mongodb-ysi-data-format}. One application of a
schema-less data model is the addition of any new key-value pair to any
instance of the document, as mongoDB supports update operations to any of the
collections. It can use tags for the purpose of annotations as described in the
taxonomy of data description. For example, in case of an oil spill in the San
Francisco Bay occurs, researchers can update the data set of a given date
range by adding tags to the relating documents. As a consequence of the use of
a schema-less data model, these changes do not affect the other data
instances, since there are no schema changes.

As described in chapter 5, the schema-less data model can provide a better
support to Data Provenance because the data model gives support to the change
of records without affecting the others. In this way, the metadata used by the
implementation clearly describes the three fundamental questions regarding the
collected data from sensor devices, as described in section
\ref{sec:keys-definition}. First, the identity of the data is given by the key
``message\underline{ }id'', in a way that it uniquely identifies the data
produced by the sensor identified by the key ``sensor.ip\underline{
}address''. In this way, the message can be tracked back to the sensor that
produced the data for different purposes such as tracking the data producer.
Second, time dimensions were used in order to track two different aspects of
the data collection: the valid time can be used to correlate observations to
the exact moment it occurred, while the transaction time can be used to verify
when the data was collected, and make decisions such as deleting the instances
of data were collected in the previous two days, etc. Last, but not least, the
use of the key ``sensor.location'' can give the exact location of the data,
since researchers can be interested in finding data based on a specific
location. Finally, the most important aspect of the schema-less data model is
that the observed data is added to the key ``observation'', as any sensor
device carries its set of attributes of key-value pairs. A good example of the
applicability of the schema-less data model is the scalability regarding
changes to the data structure. For instance, if an existing sensor device is
flushed with a new version of the firmware\footnote{The first set of machine
instructions to run on the hardware after the application of power stored
permanently in PROM or ROM or semi-permanently in EPROM}, changing the format
of data types or adding new values, the following instances of collected data
may be different to the previous one by simply adding the new additions of
data. For this reason, the DSP Data Persistence gives administrators the
ability to choose which fields to persist by declaring it using a bootstrap
message for the component, as shown in the parameter ``YSI\underline{
}DESIRED\underline{ }PROPERTIES'' of Listing
\ref{file:dsp-data-persistence-bootstrap.xml}.

The centralized query processing used by NetBEAMS is a result of the
infrastructure used by NetBEAMS, as well as the purpose of data and the
location of the collected data. As a direct result of a schema-less data
model, the centralized query processing is easily developed using any of the
available programming languages provided by mongoDB. Considering the fact that
programming languages are easier abstractions to researchers without computer
science background \cite{sn-programming-language}, mongoDB's approach to data
access through API can be seen as an easier task to manage data. As it is shown
in Listing \ref{file:experiment-query-scenarios}, the scenarios designed to
evaluate the implementation are written in Javascript, since it is the script
language by the mongoDB Shell. Therefore, it attends the requirement of not
using the SQL language as the main query language of the data, giving a better
data access abstraction similar to programming languages for different users.

One of the most important features of a data persistence for sensor
networks is the replacement of the SQL by the use of programming
languages, as they tend to be easier abstraction to users without the
background in database systems \cite{sn-programming-language}. In this way,
mongoDB covers the requirements of providing data access to the collected
sensor data in different languages. For example, the data collected by the DSP
Data Persistence component is written in Java, while the implementation of the
experiments scenarios for querying the collected data is given in Javascript. In
this way, researchers of the RTC may choose one of the enumarated languages
APIs provided by the mongoDB to access the stored data from NetBEAMS.

Finally, the implementation supported by mongoDB during the experiments was
following a single dedicated server virtualization using VirtualBox 
\cite{virtualization}.

@@

Overall, the implementation of the DSP Data Persistence component meets
the specification of the use cases defined in section \ref{sec:exp-scenarios}.
The implementation of mongoDB using a single server environment can give
support to the current workload and infrastructure from the use of NetBEAMS to
collect data from the sensors devices from SF-BEAMS. From the perspective of
disk-space utilization and given that aproximaly 1.3 Gigabytes of disk space
were used to save one year-worth of collected data equivalent to five YSI
devices. According to the mongoDB documentation \ref{mongodb}, the claimed
amount of disk space is related to how mongoDB uses a binary representation
of the JSON documents, which are composed only of strings, together with the
definition of the indexes also represent in JSON. As a consequence, the size of
the keys defined for the YSI sonde data defines the amount of disk space used.
For this reason, different experiments were conducted with different
definitions of the keys of the documents
approach when compared the the RTC's OPEnDAP format shown in Listing
\ref{file:rtc-ysi-opendap}. Then, the second specification used short names
for the keys. In this way, the amount of storage space utilized by for the
different workloads define in section \ref{sec:workload} can be summarized as
follows

The performance to insert data into the database system using the Java driver
is documented by mongoDB as the fastest approach, providing persistence to
approximaly 25,091 documents per minute on an indexed database. In contrast,
when indexes are not used, more then 169,508 documents could be inserted per
minute. Although the latter approach supports more ducuments than the former,
the trade off between insertion and selection time is relevant due to the
activities of retrieval for documents for specific periods of time. Therefore,
the use of indexes might speed the retrieval time considerably. Considering
this number of insertions to the database the number of sensors producing data
and sending to the data sink, mongoDB and NetBEAMS implementation can
virtually support thousands sensor devices.

Disk utilization is one important characteristic of the selected database
system. However, the retrieval of the collected data from sensor devices is
a much more important requirement because most of the utilization of the data
persistence of a sensor network is regarding the search for the properties on
the collected data. After analyzing the response times for the scenarios
described in section \ref{sec:exp-scenarios}, it is clear that the use of
indexes improves the overall execution of the queries, specially because of the
amount of data collected over the years. Furthermore, the response times
described in this scenario is directly related to the use of the External
Storage strategy. On the other hand, improvements over the response time can be
achieved through the implementation of a Data-Centric approach on a Distributed
Database environment, or clusters. In this approach, the inserted data is
stored in different servers based on a partition key. Given the fact that less
data is concentrated in each partition, the centralized query mechanism used is
faster, and consequently improves the performance as described in the Data
Location taxonomy in chapter 3.

When it comes to re-using the data collected by the DSP Data Persistence,
different approaches were available. First, the use of the export capabilities
to a common-delimited values was performed \ref{file:mongodb-export-command},
taking approximaly 3.3 minutes to export one year worth of data for 5 YSI
devices without indexing. Second, the use of the client shell gives a
centralized access to either the single server. However, in order to better
provide access of data to users without skills in database systems, web
applications can be built on top of the REST Web Services APIs \cite{http-rest}
provided by mongoDB. One such example of using the REST API is the adaptation
of Apache Futon 4 mongoDB, as shown in Figure
\ref{fig:view-collections-instance-browser-futondb}, where users can navigate
through the collections of the collected sensor network data categorized by the
collection. Therefore, RTC staff technicians could take advantage of this tool
to view and navigate over the collected data from the sensor devices.

@@

Supposing an environmental accident such as the San Francisco oil
  spill happened again \cite{sfbay-oilspill2009}, changing the sampling
  frequency drastically to different values.

\section{Difficulties Experienced}
\label{sec:experiments-difficulties}
As the development of computing system, the inception of the experiments
described in this chapter had different moments of Software Engineering.
Particularly speaking about the learning curve, different situations
contributed to the success of the first experiments. Since the technoogy is
new, it brought new challenges regarding the usability and user experience.
Since the technology is relatively new and still under development, many
different bugs were could be reproduced by random during the execution of the
experiments, and therefore, sent to the open-source community through use 
different available communication channels. First, the fastest way to solve an
issue was through the mongoDB's IRC channel, located at the Freenode server
(www.freenode.org, channel mongodb). Second, the use of the users group, located
at http://groups.google.com/group/mongodb-user also helped solving problems.

Physical memory is limited, for this reason, it may not be enough to run the
experiments for total workloads. In this case, the Java Virtual Memory was
totally consumed by one of the revisions of the experiments setup. Huge data
loads require different approach to manage data in memory such as creating
object pools as temporary cache for objects whose state are similar to others.

While performing the experiments, the 32bit machine reached its limitation of
2GB of physical data. With know about that, the environment needed to be
updagred for mongoDB requires a 64-bit operating system when the amount of
stored data reaches 2 Gigabytes. For this reason, the scenarios with indexes
used a dedicated virtualized server using Ubuntu Linux 9.04 64 bits with 1 Gb
of memory.

>>>>

The simulation of both the single and distributed system used both a Mac OS 10
Snow Leopard 64bit, with a total of 8GB of physical memory. However, in order
to simulate different shards in different machines, the use of a VirtualBox
\cite{virtualization} provided the use of guest operating systems to run and
act as the mongoDB shards. In this way, two different versions of the Ubuntu
Linux were used: one 8.04 32bits and another 9.04 64bits. The configuration
used for the single server only uses the process ``mongod'' to be started,
while the clustered setup must be enabled in the database, as shown in
Listing \ref{file:experiment-setup-mongoshards}, Listings
\ref{file:setup-os-shards-step1} and \ref{file:setup-os-shards-step2}.

@@

The first set of measurements collected were the ones regarding the use case
R1, shown in Listing \ref{file:dsp-mongo-service}, defined as the insertion of
instances of the collected data from the sensor devices using Java. The
document follows the specification of the keys designed in section
\ref{sec:dsp-persistence-data-model}. The use of the mongoDB Java driver, which
provides the methods to access the database system and the collection. In this
way, the construction of the keys and respective values were done using a
Hash-like object ``BasicDBObject'' such as the ones shown in the method
``buildKeySegment''. Similarly, the creation of database indexes is performed
using the same approach, as shown in the Java method ``setupMetadataIndexes''.
Finally, the insertion of the documents is performed on the service method
call ``insertPersistentUnitMessageContents'', which delimits the bulk insert
using the transaction management mechanism over the collection. In this way,
different insert times were collected from the logs of the different
experiments as summarized on Table \ref{tab:experiment-insert-avarage}.

@@

\begin{table}
    \begin{center}
        \begin{tabular}{|c|c|c|c|}\hline
        \textbf{Scenario Type} & \textbf{Language} & \textbf{Method Call} & \textbf{Use Case}\\\hline 
        \textbf{Insert} & Java & db.SondeDataContainer.insert() & R0\\\hline
        \textbf{Retrive} & Javascript & db.SondeDataContainer.find() & R1, R2, R3\\\hline 
        \textbf{Update} & Javascript & db.SondeDataContainer.update() & U1\\\hline 
        \textbf{Delete} & Javascript & db.SondeDataContainer.remove() & D1\\\hline
        \textbf{Export} & Python & db.SondeDataContainer.find() & R4\\\hline
        \end{tabular}
        \caption{mongoDB method calls used for the use cases}
    \end{center}
    \label{tab:experiment-scenarios-mongo-methods}
\end{table}

@@

\subsection{User Experience}
\label{sec:experiments-user-experimence}

In general, the use experience of mongoDB is related to the use of programming
languages, since most of the methods to create, access and modify the database
is performed using the programming language abstraction of objects and method
calls. The creation of data was the first contact with mongoDB in terms
of data creation in Java. Later, Different methods of data access were used
after the randomly generated data was inserted into mongoDB: using the shell
commands mongo and ``mongoexport'', command-line execution developed in
Python, through the REST Web Services API, and using a web application,
written in the Ruby Programming Language \cite{ruby}, providing the view over
the collected data using a web browser. More details about the execution of
these commands during the experiments is shown in section
\ref{sec:mongodb-user-experience}.

@@

While the access to the database using the shell command gives full access to 
the CRUD operations, the access to the data was also performed using the REST
Web Services API, as shown in Listing \ref{cmd:mongo-rest-request}. As one can
see, the IP address and the default port number of the mongoDB REST middleware
are provided. Then, the REST operation uses the name of the database and
the collection to correlate with the ones in the database system. In this way,
querying the collections from mongoDB can be performed using the HTTP Request
Parameter variables, by adding the prefix ``filter\underline{ }'' before the
name of the key of the document. The given example shows the query performed in the
collected data for the key ``observation.Conductivity'' with the value
``104.5''. Similarly, the HTTP Request variable ``limit'' is used to reduce
the number of documents returned by the query. Finally, the HTTP Response from
mongoDB returns the result using the JSON format, similar to the ones printed
in the command-shell.

@@

The actual execution of the solution was performed using the demo Mouse
Actions, which consists of a software sensor that perceives any mouse action in
a window area. The experiment consisted of moving the mouse around a white
window, where each action is captured and temporarily saved until a threshold is
reached, making the component to send the messages to the persistence layer, so
that it can save on mongoDB. Figure \ref{fig:experiment-mouseactions-dsp} shows
a screenshot of the moment when the message is sent to mongoDB, being verified
through the HTTP REST Interface using a web browser.

@@

The implementation of the data persistence for NetBEAMS follows the
specifications explained in chapter 4. The purpose of sensor network for
NetBEAMS is to suport data archival from any collected data, and for this
reason, mongoDB was selected after an empirical analysis described in chapter
5. In this way, the requirements described in chapter 4 were expressed as
examples of use cases in section \ref{sec:exp-scenarios}. This chapter
discusses the experiments in different angles: analysis of the taxonomies, the
infrastructure and the use cases evaluation.

@@

Any database system covers the purpose of the sensor data for NetBEAMS, which
is Data Archival, along with the type of storage for the data as an External
Storage. After selecting mongoDB, the development of the persistence layer met
the requirements of archiving data for different types of sensor devices from
the DSP Platform such as the YSI Data Handler and the Mouse Actions demo data.
As it is shown in section \ref{sec:exp-measurements}, the collected data is
archived in files using the BSON format \cite{bson}, a binary representation
used to store instances of the document described in Listing
\ref{file:mongodb-ysi-data-format}. Along with the purpose of Data Archival,
the use of mongoDB also met the specification of an External Storage, since the
data persisted in the file system, as well as Data-Centric for running mongoDB
with different  distributed mongo shards (cluster nodes).

SF-BEAMS uses Tabular Data model, Schema-less data model that represents the
collected data using comma-delimited files, without a database system to manage
those files, just the OPEnDAP middleware. In contrast, the data persistence
proposed by this work proposes mongoDB as a schema-less database system to
manage the collected data collected using NetBEAMS. As described in chapter 5,
the schema-less data model can provide a better support to Data Provenance
because the data model gives support to the change of records without affecting
the others. In this way, the metadata used by the implementation clearly
describes the three fundamental questions regarding the collected data from
sensor devices, as described in section \ref{sec:dsp-persistence-data-model}.
First, the identity of the data is given by the key ``message\underline{
}id'', in a way that it uniquely identifies the data produced by the sensor
identified by the key ``sensor.ip\underline{ }address''. In this way, the
message can be tracked back to the sensor that produced the data for different
purposes such as tracking the data producer. Second, time dimensions were used
in order to track two different aspects of the data collection: the valid time
can be used to correlate observations to the exact moment it occurred, while
the transaction time can be used to verify when the data was collected, and
make decisions such as deleting the instances of data were collected in the
previous two days, etc. Last, but not least, the use of the key
``sensor.location'' can give the exact location of the data, since researchers
can be interested in finding data based on a specific location. Finally, the
most important aspect of the schema-less data model is that the observed data
is added to the key ``observation'', as any sensor device carries its set of
attributes of key-value pairs. A good example of the applicability of the
schema-less model is the scalability regarding changes to the data structure.
For instance, if an existing sensor device is flushed with a new version of
the firmware\footnote{Primary machine instructions that run in the hardware},
changing the format of data types or adding new values, the following instances
of collected data may be different to the previous one by simply adding the
new additions of data. For this reason, the DSP Data Persistence gives
administrators the ability to choose which fields to persist by declaring it
using a bootstrap message for the component, as shown in the parameter
``YSI\underline{ }DESIRED\underline{ }PROPERTIES'' of Listing
\ref{file:dsp-data-persistence-bootstrap.xml}.

In order to access the collected data, the implementation used different APIs
provided by the mongoDB community. The Centralized Query Mechanism was
satisfied by the use of those APIs, which is also fultill the requirements
defined by NetBEAMS. For example, the access of the database using Python or
Java were based on the needs and the use of the languages for different
purposes. In this way, the implementation of the persistence model captures
the instances of any known data collected from NetBEAMS and saves it using the
document pattern as shown in Listing \ref{file:mongodb-ysi-data-format}. Any
programming language used was capable of accessing the same collected data by
their own.

In the end, the implementation of the data persistence for NetBEAMS was
developed in both Single and Distributed Systems. The use of regular single
server execution provided enough computational power for a compatible volume of
data when compared to the one produced from the SF-BEAMS sensor network. In
this way, the implementation suggested by this work fulfill the need of a
persistence system for a Small volume of data being created at the same type
in a Single Server. Furthermore, mongoDB also provided support for Distributed
System, when the use of databse shards, in forms of different mongoDB nodes,
were used as a clustered environment.

@@

Overall, the implementation of the DSP Data Persistence component meets the
specification of the use cases defined in section \ref{sec:exp-scenarios}.
From the perspective of disk-space utilization for the compatible volume of
data for 5 five YSI devices is that it requires a small amount of disk space.
According to the mongoDB documentation \cite{mongodb}, the claimed amount of
disk space is related to how mongoDB uses a binary representation of the JSON
documents, which are composed only of strings, together with the definition of
the indexes. As a consequence, the size of the keys defined for the YSI sonde
data suggests additional disk space were used for the storage.

The performance to insert data into the database system using the Java driver
is documented by mongoDB as the fastest approach, providing persistence
thousands of documents per second. This capacity is more than necessary for
NetBEAMS, which can use the Single System environment. On the other hand,
improvements over the response time can be achieved through the implementation
of a Data-Centric approach on a Distributed Database environment, or clusters.
In this approach, the inserted data is stored in different servers based on a
partition key. Given the fact that less data is concentrated in each
partition, the centralized query mechanism used is faster, and consequently
improves the performance as described in the Data Location taxonomy in chapter
3. In this way, the infrastructural performace during the inserts suggests that
persisting data collected from the NetBEAMS sensors using a clustered
environment is an option when the single server capacity is reached.

Different problems and limitations were faced during the execution of
mongoDB, which were relate to the infrastructure used: the limitations
regarding disk storage in 32bit operating systems and very slow
performance when using virtualized servers. During the execution of the
experiments, mongoDB used memory accordingly, without taking control of the
operating system's resources. However, when mongoDB is used in a 32bit server
architecture, there are limitations on the disk space used to store
collections. In this way, after reaching 2GB of collected data during many
different runs, mongoDB crashed and did not allow the use of the collected
data. For this reason, the setup of a new VirtualBox using Ubuntu Linux 64bits
as a guest to a Mac OS X 64bits was prepared and used throughout the completion
of this work's experiments. On the other hand, the performance was drastically
decreased using mongoDB in virtualized servers due to the fact of dynamic
partition chosen. The performance shell application ``iostat'' was used to
verify the behavior, and the use of mongoDB was switched to the Host operating
system.

@@

\cite{sn-programming-language} suggested that programming that programming
languages are easier abstractions to researchers that use collected data from
sensor devices without background in computer science or database systems. The
authors have suggested the implementation of a programming language for this
group of biologists, for instance. However, this suggestion has good
intentions, but limits the users with a single mechanism to access the
collected data from sensors. For this reason, this dissertation proposes the
use of key-value pair databases, such as mongoDB, to provide transparent access
the the data through the abstraction of any programming language that has a
driver written for mongoDB. As one can see during the user experience section,
different programming languages were used to access the collected data, a novel
mechanism not yet reported in the sensor networks community.

The development of the scenarios were clearly written in different ways, such
as the ones defined in Listing \ref{file:experiment-scenarios}, which
implements the use cases for data retrieval, update and deletion. The main
advantage of this approach is that users do not use the SQL language to extract
data from the database. Rather, the use of the method calls defined for each
operation, as well as the concept of dynamic binding and objects, gives simple,
yet powerful syntax to access data. For example, based on the description
of the collected data by keys and values shown in Listing
\ref{file:mongodb-ysi-data-format}, different use cases can evaluate documents
in this structure. Usually researchers may search the collected data by a given
range of dates, for their regular activities of data analysis. The scenario
implemented by R1 in the abovecited listing shows it clearly. Similarly, the
search for data in a given sensor device is expressed by the search for the IP
address for them on the use cased defined by R2. Finally, the search for the
raw data obtained by the sensor devices is used from the key ``observation'',
as shown by the R3 use case implementation.

Accompanying that non-predictable events might occur during the data
collection process of sensor devices, and consequently, influence in their
values, the support to annotations is a feature that can be used to describe
the data. mongoDB indirectly supports the use of annotation because of its free
method of updating documents and adding new keys to any of them without
requiring design changes. Supposing an environmental accident happens, such as
the San Francisco oil spill of October 2009 \cite{sfbay-oilspill2009}, the
values of the collected data, and consequently data analysis, be influenced by
different ranges of values. In this way, the implementation of the scenario U1
provided the implementation of the addition of a new key called ``tag'', with
the annotation ``oil spill''. This new key, after properly indexed, provides
another option to meaningfully search the collected data from the sensor
devices. Last, but not least, old data can be deleted as expressed by D1, as an
operation supported by any database system.

One common approach during research is data sharing. Researchers can choose
the use of spreadsheets to analyze collected data with simpler tools than
database systems or programming language. For this reason, OPEnDAP is used in
the community to share data using a comma-delited format. In this way, the
implementation of the use cases for R4 were developed in two different ways.
First, the use of the export capabilities to a common-delimited values was
performed \ref{file:mongodb-export-command}, taking approximaly 3.3 minutes to
export one year worth of data for 5 YSI devices without indexing. Second, the
implementation of a custom export utility using Python, as shown in Listing
\ref{file:experiment-export-python} enabled the creation of the comma-delimited
files with the same format as the ones used by OPEnDAP. The comparison can be
done by analyzing the output of the OPEnDAP request from SF-BEAMS on Listing
\ref{file:rtc-ysi-opendap}, and the output provided from the script written in
Python, show in Listing \ref{file:experiment-export-python-results}.
Furthermore, considering research teams with skills in Web Services, an easy
ubiquos integration with mongoDB can be provided by using the REST Web Services
APIs provided by mongoDB, still in alpha version. A practical application of
this API was the created Apache Futon 4 mongoDB web application, used to give
user friendlier access to the collected data as shown in 
\ref{fig:view-collections-instance-browser-futondb}.

As discussed in chapter 5, the differences between a traditional database
system, which uses the relational data model, and a document-oriented database
system were clearly noticed during the development of this work. As mentioned
in the section \ref{sec:experiments-user-experimence}, the user experience was
mostly related to the definition of the keys that describe the sensor device,
as discussed during the design of the solution in chapter 6. On the contraty,
the abstraction of objects and collections during the data access using
different programming languages. Therefore, based on the use of a different
type of data model in the scope of data persistence for sensor networks, the
taxonomies defined in chapter 3 could be updated to the one shown in Figure
\ref{fig:taxonomy-data-model-modified}, including the Key-Value Pair Data
Model.

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.5]{../diagrams/taxonomy-data-model-modified}
  \caption{Data Model Taxonomy}
  \label{fig:taxonomy-data-model-modified}
\end{figure}

\subsection{Difficulties Experienced}
\label{sec:experiments-difficulties}

The inception of the experiments described in by this work moments of Software
Engineering. Particularly speaking about the learning curve, different
situations contributed to the success of the first experiments. Since the
technoogy is new, it brought new challenges regarding the usability and user
experience. The technology is relatively new and still under development,
which many different bugs were be reproduced by random during the execution of
the experiments, and therefore, reports were sent to the open-source community
through use different available communication channels. First, the fastest way
to solve an issue was through the mongoDB's IRC channel located at the Freenode
server (www.freenode.org, channel mongodb). Second, the use of the users
group, located at http://groups.google.com/group/mongodb-user also helped
solving problems.

Physical memory is limited and, for this reason, it may not be enough to run
the experiments for total workloads. In this case, the Java Virtual Memory was
totally consumed by one of the revisions of the experiments setup. Therefore, 
it was learned that large amount of data loads require different approach to
manage data in memory accordingly, such as creating different object pools as
temporary cache for objects whose state are similar to others.

Also related to the difficulties of this work, the fact that mongoDB is not yet
stable makes the implementation a bit harder. The main mongod process is on its
1.1.4 version as the time this line is writte, while the shards support is
still in alpha version. For instance, different work-arounds were performed to
have shards experimented as described, but the execution was not successful.

To sum up, this chapter described the experimental results of the
implementation of a data persistence component for NetBEAMS, showing the
experiment results, the measurement results and finally the discussion about
the solution in general. The following section brings the conclusions and
future works for this dissertation.

### 2009-12-08 docs/marcello-dissertation/thesis/chapter9-conclusions-and-future-works.tex
First Draft of the dissertation peer reviewed with Kleber.
git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@680 559edde2-a4d4-44e0-96fc-4035e40d7396

\chapter{Conclusions}

This work presented a collection of taxonomies related to data persistence for
sensor networks using different approaches ranging from data models and server
infrastructure organization. The main result of this work was that persistence
for collected data from sensors are better maintained using schema-less data
models, since it provides offers an approach to deal with unanticipated data
types from the sensor devices.

As seen in the state of the art, different approaches can be done in order to
evaluate data persistence for nsensor networks. T

There are many important features covered in this work, including the
importance of the sensor networks rely in how users have access to the
collected data. In order to provide access to the collected data, one must
first assess the different types of infrastructural characteristics of the
studied sensor network. In general, the selection of a technology that
provides such persistence capabilities can be part of a process of analysis of
taxonomies proposed by this work. However, this work proposes a data model not
yet used in the sensor network community and, thus, represents an important
contribution for the area.

Taking into account the software architecture provided by NetBEAMS, as well as
the requirements for data collection from the RTC research group, this work
shows a novice way to persist data collected from sensor devices by using a
schema-less data model, whose query capabilities are based upon programming
languages, in contrast to the use of the Structured Query Language, SQL. The
experiment results showed that the implementation of a persistence layer with
mongoDB can be used to persist data at its maximum rate of one sample per
minute. However, the evaluation of mongoDB was not used to implement a
Data-Centric Storage approach, which can decreases the amount of time to search
data since the process uses parallel search.

All in all, the DSP Data Persistence component implemented for the NetBEAMS
solution to SF-BEAMS can be used in any network server. The solution for data
persistence is a novice way to save data into the database, provided by
mongoDB's ability to cope with the uncertainty of any properties collected from
any sensor specification. Furthermore, the solution also provides a good
support to the provenance-related data as the schema-less data model gives the
ability to .

\chapter{Future Works}

0. Todo list for this work
1. Database System Problems
2. Software Engineering Problems
3. Applications
 - plot graphs of collected,

There are many different directions for future works. First, the
implementation of a Data-Centric Storage approach can help the system scale
horizontaly by using the mongoDB's database shards. In this way, this approach
can reduce the retrieval time by focusing on centralized selection of data over
a given database shard. Moreover, the overall search through the collected data
attributes can be drastically decreased by parallel search over the database
shards using the MapReduce \cite{map-reduce} tools. At this time of
development, the MapReduce API offered by mongoDB is on its alpha version, and
therefore, presenting lots of bugs. Figure
\ref{fig:future-works-data-centric-map-reduce} shows an architectural idea of
the deployment of a Data-Centric storage approach. When searching for specific
document attributes, the process is sent in parallel to each of the database
shards, and the result is collected by the cluster head. After calculating the
final result, mongoDB server returns the relevant result to the user.

>>>>

\chapter{Conclusions and Future Works}

There were important important remarks covered in this work, including the
importance that sensor networks are becoming ubiquitous, having their produced
data available online and other types of media. In this way, in order to
provide access to the collected data, one must first assess the different
types of infrastructural characteristics of the studied sensor networks. In
general, the selection of a technology that provides such a persistence
solution capabilities can  part of a process of analysis of taxonomies
proposed by this work, characterized as the first contribution of this
disseration.

In my own opinion, based on the findings of the taxonomies, together with
the empirical analysis of the technologies, this work revealed a novel approach
to provide persistence and access to collected data from sensor devices using
the Key-Value-Pair data model. Besides the contribution of a new component
component for the Data Sensor Platform, the evidence that the system improves
productivity was confirmed during the execution of the experiments, writing
less source-code to obtain better functionalities when compared to other
solutions. Similarly, improved work suggests that the abstraction of
programming language should be used with research groups whose primary
expertise is not in Computer Science or Database Systems.

All in all, the DSP Data Persistence component implemented for the NetBEAMS
solution to SF-BEAMS can be used in any network server. The solution for data
persistence is a novice way to save data into the database, provided by
mongoDB's ability to cope with the uncertainty of any properties collected from
any sensor specification.

As far as future works, there are many different directions for future works.
First, the first major release of mongoDB with complete support to shards is
scheduled for 2010. An important mechanism of mongoDB is the fact that it can
manage different mongoDB servers spread as a cluster, as it gives support to
the Data-Centric Storage approach. In this way, this approach can reduce the
retrieval time by focusing on centralized selection of data over a given
database shard. Moreover, the overall search through the collected data
attributes can be drastically decreased by parallel search over the database
shards using the MapReduce \cite{map-reduce} tools. At the time of development
of the experiments, the MapReduce API from mongoDB was also in the alpha
version, and therefore, presenting lots of bugs. one suggestion for the
implementation of such functionality is based in Figure
\ref{fig:future-works-data-centric-map-reduce}, which shows an architectural
idea of the deployment of a Data-Centric storage approach. When searching for
specific document attributes, the process is sent in parallel to each of the
database shards, and the result is collected by the cluster head. After
calculating the final result, mongoDB server returns the relevant result to
the user. I was personally expecting to implement the scario of data retrieval
of collected daa.
